Lucid: Lifting-based Uncertain Control Invariant Dynamics

### 2025-06-16 docs/paper/main.tex
feat: add case studies

\title{Delta-Complete Linear Programming}

@@

Lucid is in implemented in C++ and it is meant to be used as a library.
It provides a set of interfaces that allow users to have an high level of control and flexibility over the verification process.
We also provide a Python wrapper, called PyLucid, to facilitate the integration of the tool into existing workflows and effortlessly leverage well-established libraries such as NumPy~\cite{numpy} and SciPy~\cite{scipy}.
Lucid is available as an open-source project on GitHub at \url{https://github.com/TendTo/lucid}.
More technical details about the implementation can be found in the documentation at \url{https://tendto.github.io/lucid/}.

@@

\subsection{Notation}
\label{sec:notation}

@@

\subsection{Control Barrier Certificates}
\label{sec:cbc}
Whilst certifying the safety of a stochastic system is generally challenging, \emph{control barrier certificates} (CBCs) and \emph{control barrier functions} (CBFs) leverage the concept of set invariance to arrive at an abstraction-free formulation. This has made CBCs/CBFs popular tools for safety verification and synthesis~\cite{prajna2006barrier,cosner2023generative}.
We briefly recall the definition of a CBC.
\begin{definition}[Control barrier certificate (CBC)]\label{def:cbc}
  A function $\B:\X\rightarrow\R_{\geq 0}$ is called a CBC of an MDP $\M=(\X,\X_0,\U,\Tr)$ with reference to an unsafe set $\X_u$, if we have
  \begin{itemize}
    \item[(a)] $\forall x_0\in \X_0\!:\,\B(x_0)\leq\eta$;
    \item[(b)] $\forall x_u\in \X_u\!:\,\B(x_u)\geq\gamma$; and
    \item[(c)] $\forall x\in\X,\exists u\in\U\!:\,\E_{\Tr}[\,\B(X^+)\mid X=x,\,U=u\,] -\B(x) \leq  c;$
  \end{itemize}
  for some constants $\gamma>\eta\geq0$ and $c\geq 0$.
\end{definition}
If one can find a CBC for a system $\M$, then, a lower bound on the probability of $\M$ being safe is given by the following proposition due to \cite{kushner1967stochastic}.
\begin{proposition}[Finite-horizon safety]
  \label{prop:reach}
  Consider an MDP $\M=(\X,\X_0,\U,\Tr)$ and a safety specification $\psi_{\text{safe}}=(\X_u,T)$. Suppose there exists a CBC $B$ w.r.t. $\X_u$ (Definition~\autoref{def:cbc}) with constants $\eta$, $\gamma$, and $c$. Then, there exists a stationary policy $\pi$ such that
  $$P^\pi_{\text{safe}}(\M)\geq 1- \frac{\eta + cT}{\gamma}.$$
\end{proposition}

@@

When verifying the safety of a system, \lucid will search for a CBC $\B$ that satisfies the conditions in Definition~\ref{def:cbc}.
If it finds it, it will return the constants $\eta$, $\gamma$, and $c$, as well as the CBC $\B$ itself, which becomes a certificate of safety for the system $\M$.
The template of the barrier function is explored later in section~\ref{sec:benchmarks}.

@@

Via Lemma~\ref{lem:trig-bound}, we obtain a finite-constraint relaxation of the SIP \eqref{eq:semiinf-prog} based on the constraint-tightening reasoning
\begin{equation*}
  \frac{1}{2}\big(\mathrm{max}_{\hat{N}}^\X + \mathrm{min}_{\hat{N}}^\X - C_{\hat{N}}\left(\mathrm{max}_{\hat{N}}^\X-\mathrm{min}_{\hat{N}}^\X\right)\big)\geq \mathrm{lb}
  \quad\Longrightarrow\quad
  \forall x\in\X: f(x) \geq \mathrm{lb},
\end{equation*}
where $\mathrm{max}_{\hat{N}}^\X:=\max_{x\in X_{\hat{N}}} \left\lbrace\phi_M(x)\T b\right\rbrace$ and $\mathrm{min}_{\hat{N}}^\X:=\min_{x\in X_{\hat{N}}} \left\lbrace\phi_M(x)\T b\right\rbrace$ for some desired lower bound $\mathrm{lb}\in\R$.
Thus, we obtain strengthened constraints of the form
\begin{align*}
                               & \frac{1}{2}\big(\mathrm{max}_{\hat{N}}^\X + \mathrm{min}_{\hat{N}}^\X - C_{\hat{N}}\left(\mathrm{max}_{\hat{N}}^\X-\mathrm{min}_{\hat{N}}^\X\right)\big) \geq \mathrm{lb}, \\
  \Leftrightarrow\hspace{10pt} &
  \mathrm{min}_{\hat{N}}^\X \geq \frac{2}{C_{\hat{N}}+1}\mathrm{lb} + \frac{C_{\hat{N}}-1}{C_{\hat{N}}+1}\mathrm{max}_{\hat{N}}^\X.
\end{align*}
For an upper bound $\forall x\in\X: f(x) \leq \mathrm{ub}$, we obtain analogous constraints of the form
\begin{equation*}
  \mathrm{max}_{\hat{N}}^\X \geq \frac{2}{C_{\hat{N}}+1}\mathrm{ub} + \frac{C_{\hat{N}}-1}{C_{\hat{N}}+1}\mathrm{min}_{\hat{N}}^\X.
\end{equation*}

@@

\section{Benchmarks}
\label{sec:benchmarks}

>>>>

\title{Lucid: Lifting-based Uncertain Control Invariant Dynamics}

@@

\paragraph{Motivation.}

We present a software tool, called \lucid, able to address the challenges of verifying the safety of black-box noisy systems.
A description that, for better or for worse, fits many modern AI-enabled systems, such as autonomous vehicles, robotics, and healthcare applications.
To ensure their safety, we use \glspl{cbc}~\cite{prajna2006barrier,cosner2023generative}, which are a powerful tools that provides formal guarantees and a verifiable witness.
The peculiar function template we chose for the \glspl{cbc} makes it easier for the model to be learned from limited data collected from the system, without the need for an explicit dynamical model.
Note that the current version of the tool does not deal with external control inputs, which we plan to address in future work.

\paragraph{Tool.}

\lucid is implemented in C++ and it is meant to be used as a library.
It provides a set of interfaces that allow users to have an high level of control and flexibility over the verification process.
We also provide a Python wrapper, called PyLucid, to facilitate the integration of the tool into existing workflows and effortlessly leverage well-established libraries such as NumPy~\cite{numpy} and SciPy~\cite{scipy}.
\lucid is available as an open-source project on GitHub at \url{https://github.com/TendTo/lucid}.
More technical details about the implementation can be found in the online documentation at \url{https://tendto.github.io/lucid/}, but at an high level, the architecture of \lucid is illustrated in~\autoref{fig:architecture}.

\input{fig/architecture}

\paragraph{Literature Background.}
\todo{Add relevant literature citations and background information here.}

% \input{fig/architecture.tex}

@@

@@

\subsection{Control Barrier Certificates}
\label{sec:cbc}
Whilst certifying the safety of a stochastic system is generally challenging, \emph{control barrier certificates} (CBCs) and \emph{control barrier functions} (CBFs) leverage the concept of set invariance to arrive at an abstraction-free formulation. This has made CBCs/CBFs popular tools for safety verification and synthesis~\cite{prajna2006barrier,cosner2023generative}.
We briefly recall the definition of a \gls{cbc}.
\begin{definition}[Control barrier certificate (CBC)]\label{def:cbc}
  A function $\B:\X\rightarrow\R_{\geq 0}$ is called a \gls{cbc} of an MDP $\M=(\X,\X_0,\U,\Tr)$ with reference to an unsafe set $\X_u$, if we have
  \begin{itemize}
    \item[(a)] $\forall x_0\in \X_0\!:\,\B(x_0)\leq\eta$;
    \item[(b)] $\forall x_u\in \X_u\!:\,\B(x_u)\geq\gamma$; and
    \item[(c)] $\forall x\in\X,\exists u\in\U\!:\,\E_{\Tr}[\,\B(X^+)\mid X=x,\,U=u\,] -\B(x) \leq  c;$
  \end{itemize}
  for some constants $\gamma>\eta\geq0$ and $c\geq 0$.
\end{definition}
If one can find a \gls{cbc} for a system $\M$, then, a lower bound on the probability of $\M$ being safe is given by the following proposition due to \cite{kushner1967stochastic}.
\begin{proposition}[Finite-horizon safety]
  \label{prop:reach}
  Consider an MDP $\M=(\X,\X_0,\U,\Tr)$ and a safety specification $\psi_{\text{safe}}=(\X_u,T)$. Suppose there exists a CBC $B$ w.r.t. $\X_u$ (Definition~\autoref{def:cbc}) with constants $\eta$, $\gamma$, and $c$. Then, there exists a stationary policy $\pi$ such that
  $$P^\pi_{\text{safe}}(\M)\geq 1- \frac{\eta + cT}{\gamma}.$$
\end{proposition}

@@

When verifying the safety of a system, \lucid will search for a \gls{cbc} $\B$ that satisfies the conditions in Definition~\ref{def:cbc}.
If it finds it, it will return the constants $\eta$, $\gamma$, and $c$, as well as the CBC $\B$ itself, which becomes a certificate of safety for the system $\M$.
The template of the barrier function is explored later in section~\ref{sec:benchmarks}.

@@

Via Lemma~\ref{lem:trig-bound}, we obtain a finite-constraint relaxation of the SIP \eqref{eq:semiinf-prog} based on the constraint-tightening reasoning
\begin{equation*}
  \frac{1}{2}\big(\mathrm{max}_{\hat{N}}^\X + \mathrm{min}_{\hat{N}}^\X - C_{\hat{N}}\left(\mathrm{max}_{\hat{N}}^\X-\mathrm{min}_{\hat{N}}^\X\right)\big)\geq \mathrm{lb}
  \quad\Longrightarrow\quad
  \forall x\in\X: f(x) \geq \mathrm{lb},
\end{equation*}
where $\mathrm{max}_{\hat{N}}^\X:=\max_{x\in X_{\hat{N}}} \left\lbrace\phi_M(x)\T b\right\rbrace$ and $\mathrm{min}_{\hat{N}}^\X:=\min_{x\in X_{\hat{N}}} \left\lbrace\phi_M(x)\T b\right\rbrace$ for some desired lower bound $\mathrm{lb}\in\R$.
Thus, we obtain strengthened constraints of the form
\begin{align*}
  \mathrm{min}_{\hat{N}}^\X \geq \frac{2}{C_{\hat{N}}+1}\mathrm{lb} + \frac{C_{\hat{N}}-1}{C_{\hat{N}}+1}\mathrm{max}_{\hat{N}}^\X.
\end{align*}
We obtain an analogous constraints for an upper bound $\forall x\in\X: f(x) \leq \mathrm{ub}$.

The finite-constraint relaxation of the SIP \eqref{eq:semiinf-prog} is what \lucid uses to synthesize the \gls{cbc} $\B$.
More precisely, we follow the steps outlined in~\autoref{fig:steps}.
While the theory gives us precise indication on how to construct the \gls{lp} which would yield a correct-by-design solution,
in practice the problem may become too conservative.
To mitigate this, we can either increase the number of lattice points $Q$, worsening the computational efficiency, 
or we may choose to adjust the coefficient $C_{\hat{N}}$, improving performance at the cost of loosening the guarantees, 
which can still be checked with the verifier, provided we have access to the system's model.

\input{fig/steps}

@@

\section{Tool Evaluation}
\label{sec:evaluation}

We evaluate \lucid on a set of benchmarks, which are described in detail in the documentation.

\paragraph{Linear System}

To showcase the basic functionality of \lucid, we consider a simple linear system with the dynamics

\begin{equation*}
  \begin{bmatrix}
    {x}_{1, t+1}
  \end{bmatrix}
  = \begin{bmatrix}
    0.5
  \end{bmatrix}
  \begin{bmatrix}
    {x}_{1, t}
  \end{bmatrix} + w_t,
\end{equation*}
where $w_t\sim\mathcal{N}(\cdotx\vert 0,0.01I)$ is a one-dimensional Gaussian noise.
We restrict the state space to the interval $\X=[-1,1]$ and define the initial set $\X_0=[-0.5,0.5]$ and the unsafe set $\X_u=[-1,-0.9]\cup[0.9,1]$.
Translating the system into the \lucid format, we obtain the following configuration file:
\lstinputlisting[language=Python,caption={\texttt{linear.py} configuration file.}]{code/linear.py}
It can then be run with the command
\begin{lstlisting}[language=bash]
python -m pylucid tests/bindings/pylucid/MinScenarioConfig.py --seed 42 --gamma 1.0 \
  --time_horizon 5 --num_samples 1000 --lambda 1e-3 --sigma_f 15.0 \ 
  --sigma_l 1.75555556 --num_frequencies 4 --plot --verify --oversample_factor 32.0
\end{lstlisting}
to produce the results shown in~\autoref{fig:linear}.
By including the \texttt{--verify} flag, also get formal guarantees that the barrier is indeed a valid \gls{cbc} for the system.
\begin{lstlisting}[language=bash]
# ...
[info] [thread 1] [-] The barrier has been verified via dReal
# ...
\end{lstlisting}

\begin{figure}[ht]
  \import{fig}{linear.pgf}
  \caption{\gls{cbc} synthesis for the linear system benchmark.
    The blue regions represent the initial set $\X_0$, while the red regions represent the unsafe set $\X_u$.
    The pink surface indicates the value of $\gamma$.
  }
  \label{fig:linear}
\end{figure}

\paragraph{Barrier 3}

The first benchmark is inspired by the example \barr from \cite{abate2021fossil}, which we extend by adding stochastic noise $w_t\sim\mathcal{N}(\cdotx\vert 0,0.01I_2)$ to arrive at the two-dimensional nonlinear stochastic dynamics
\begin{equation*}
  \begin{bmatrix}
    {x}_{1, t+1} \\
    {x}_{2, t+1}
  \end{bmatrix}
  = \begin{bmatrix}
    {x}_{2, t} \\
    \frac{1}{3} {x}^3_{1, t} - {x}_{1,t} - {x}_{2,t}
  \end{bmatrix} + w_t.
\end{equation*}
We illustrate the setup in Figure~\ref{fig:CSBarr3}.

The goal is to compute the probability that the system initialized in $[{x}_{1,0},{x}_{2,0}]\T\in \X_0$ (blue regions) does not enter the unsafe regions $\X_u$ (in red) within $T=10$ time steps.

\begin{figure}[ht]
  \import{fig}{barrier3.pgf}
  \caption{\gls{cbc} synthesis for the \barr benchmark.
    The blue regions represent the initial set $\X_0$, while the red regions represent the unsafe set $\X_u$.
    The pink surface indicates the value of $\gamma$.
  }
  \label{fig:CSBarr3}
\end{figure}
